Learning constraints in spreadsheets and tabular data

### 2016-09-23 Paper/ECAI/tabular.tex
abstract, intro and rel. work pass

\begin{abstract}
Spreadsheets, CSV (comma separated value) files and other tabular data representations are in wide use today.
However, writing, maintaining and identifying good formulas for tabular data and spreadsheets can be time consuming and error-prone.
In this work, we investigate the automatic learning of constraints (formulas and relations) in raw tabular data in an unsupervised way.
We represent common spreadsheet formulas and relations through predicates and expressions whose arguments must satisfy the inherent properties of the formula/relation. The challenge is to automatically infer the set of constraints that is present in the data, without user information or labeled examples. 
We propose a two-stage generate and test method where the first stage uses constraint solving techniques to efficiently reduce the number of candidates in a generic way, based on the predicate signatures. 
Our approach takes inspiration from inductive logic programming, constraint learning and constraint satisfaction.
We show that we are able to accurately discover constraints in spreadsheets from various sources.
\end{abstract}

@@

% Moved up
The question that we ask in this paper is: is it possible to learn the set of constraints present in a spreadsheet, from just flat tabular spreadsheet data? The main challenge is the number of possible constraints and combinations of rows and columns that need to be tried as input to the constraints.
To answer this challenge we propose a general-purpose method and system, named \sname (from: Tabular Constraint Learner, pronounced ``tackle''), for discovering row-wise and column-wise constraints.
%
% contribs
Our contributions are as follows:
\begin{itemize}
\item we define the tabular constraint learning problem, where the goal is to constraints that range over entire rows or columns in an unsupervised way;
\item we propose an effective 2-stage generate and test method where the first stage reasons only over properties of contiguous blocks of rows/columns, and the second stage continues to investigate individual rows and columns and their content;
\item furthermore, in the first stage we use a constraint solver to efficiently enumerate all combinations of maximally contiguous blocks compatible with the candidate constraint's argument; %, using backtracking search and forward checking;
\item experiments on different publicly available spreadsheets show that the system is able to extract constraints with high precision and recall. 
\end{itemize}

@@

This paper is organized as follows.
Section \ref{sec:related_work} presents a detailed discussion of the related work. Section \ref{sec:formalization} introduces concepts relevant to our approach and our problem statement.  Section \ref{sec:approach} presents our approach. Section \ref{sec:evaluation} presents the evaluation of the approach. Section \ref{sec:applications} shows how our system can be used for applications.  Section \ref{sec:conclusions} provides conclusions.

@@

First, it borrows techniques from logical and relational learning \cite{luc_book}, as it finds
constraints in multiple relations (or tables). Unlike typical logical and relational learning
approaches, it focuses on data in spreadsheets and constraints on columns or rows in the sheets.
Furthermore, it derives a set of simple ``atomic'' constraints rather than a set of rules that each consist of a conjunction of literals as in clausal discovery \cite{claudien,lallouet}. Nevertheless, several ideas from logical and relational learning
proved to be useful, such as the use of a canonical form, a refinement graph, and pruning for redundancy. It also connects with the logical generalization of given formulas in a spreadsheet \cite{Isakowitz}, where one abstracts present formulas by introducing parameters in the place of particular cell references. The key difference here is that we discover constraints (formulas) in the data, we do not have or use any prior information on the formulas in a spreadsheet.

Second, there exist a number of algorithms in the constraint programming community that induce constraints from one or more examples and questions. Two well-known approaches include ModelSeeker \cite{modelseeker} and Quacq \cite{Quacq}.
The former starts from a single example in the form of a vector of values and then exhaustively looks for the most specific constraints from a large constraint library that hold in the vector. To this aim, it cleverly enumerates different de-vectorized tables; for instance, if the initial vector was of dimension $1 \times 20$, it would consider rearrangements of size $2 \times 10$, $4 \times 5$, $5 \times 4$, $\dots$ and Modelseeker would then look for both column, row  and even diagonal constraints. Key differences are that we assume given spreadsheets with \textit{multiple} tables; and that ModelSeeker's constraints are restricted to integer values and often typical for constraint programming community, e.g. aimed at combinatorial optimization and scheduling. Conacq \cite{Conacq} acquires a conjunction of constraints using techniques inspired on Mitchell's version spaces to process examples; its successor Quacq uses active learning to query the user regarding satisfiability. Each of the examples/queries is processed for each constraint, hence an important issue is how fast the algorithm converges to the entire solution set (e.g. for Sudoku).
The focus is again on integer variables and typical combinatorial optimisation constraints. In a similar spirit the field of equation discovery \cite{equation_discovery} focuses, guided by  heuristics based on the structure of physical equations and unit measures, on the discovery of physical formulas over numbers and times series.
However, they focus specifically on noisy physical data and formulas, e.g. a single variable can take values from a series of measurements to get a regression parameter estimate or they apply specific data transformation such as numerical calculation of time and partial derivatives, i.e. the method works with approximations; the equation structures are heavily biased towards parameter estimation, while we focus on discovery of different types of spreadsheet constraints which is an exact constraint satisfaction problem.
\sergey{rewritten, Samuel, does it sound fine now?}
\tias{Last sentence too vague, what is specific about physical data? btw, it is supervised, e.g. input-output and we are general? Also, it is 'one' conjunction of functions, e.g. one equation, or not? Please rewrite more precisely}
%As the conjunction for typical constraint programming problems (such as Sudoku) can be quite large, a key problem is to cope with redundant constraints. As Modelseeker, these systems focus on the types of constraints in combinatorial optimization but unlike ModelSeeker and our approach, they do not assume the variables are organized in tabular form.

@@

    This example combines several Excel sheets based on exercise sessions in the book ``MS Excel 2010''~\cite{excel_book}.

@@

  \caption{Constraints learned for the tables above, except for \textit{ALLDIFFERENT} (18), \textit{PERMUTATION} (2) and \textit{FOREIGNKEY} (5) constraints.

@@

\subsubsection{Constraints} \label{sec:which_cons}
The \samuel{count} constraints that are currently supported in our system are shown in Table~\ref{table:constraints}.
We included most formulas that we encountered in tutorial spreadsheets including the popular \textit{SUM} and \textit{LOOKUP} constraints\footnote{\href{https://support.office.com/en-us/article/Excel-functions-by-category-5F91F4E9-7B42-46D2-9BD1-63F26A86C0EB}{https://support.office.com/en-us/article/Excel-functions-by-category-5F91F4E9-7B42-46D2-9BD1-63F26A86C0EB}}.
We also added three structural constraints (\textit{ALLDIFFERENT}, \textit{PERMUTATION} and \textit{FOREIGNKEY}) so that they can be used in the signature of other constraint templates; they are also popular in constraint satisfaction~\cite{modelseeker}.

@@

\paragraph{Experimental setup}
\label{sec:evalualtion:method}
All spreadsheets used in the experiments are in CSV format, and for each we have ran the table extraction tool once (using a combination of the rather primitive automatic detection + visual verification that the tables are meaningful (e.g. merge split-up tables because of None values, or remove headers of all-textual tables).
\samuel{Should we not say just with the visual tool?}
 \\ % split of weird sentence
Blocks are detected automatically using our block detection algorithm (Section~\ref{sec:make_groups}) unless a spreadsheet contains ambiguous \textit{None} values.
For those ambiguous cases the blocks are specified manually. \samuel{I have yet to make this more precise}

% Group definitions are only specified if the spreadsheet contains ambiguous \textit{None} values. \tias{What does this mean, 'group definitons'?? Is this 'block detection' (Sec 4.2), if so, how do we explain this process??}

All spreadsheets also have a set of ground-truth of constraints, we call these the \textbf{intended} constraints, that are expected to be (re-)discovered.
These were determined by us using either the formulas of the original sheet (if present), or inferring them based on, for example, the header names in the sheet.
In the following experiments, we focus on spreadsheets specifically and hence only include \textit{functional} constraints, constraints that can be expressed as formulas. We ignoring structural constraints such as \textit{ALLDIFFERENT}, \textit{FOREIGNKEY} and \textit{PERMUTATION}. All constraints are stored in their canonical form. 5 spreadsheets contained rows or columns that were entirely identical, and any intended constraint involving one such vector was also added for the equivalent vectors.

Section~\ref{sec:which_cons} discusses the 32 constraints currently supported by \sname \tias{It does, right? Should also contain '32' number}\sam{I will recount at the end}. Some of the intended constraints are currently outside of the scope of \sname however, more specifically nested mathematical or nested logical formulas. We denote by \textbf{supported} constraints the subset of intended constraints that the system can find in theory.

@@

Hence, for this example we achieve a recall of 100\% (all intended constraints are discovered) and a precision of $12/18 = 67\%$. All intended constraints are supported. The recall is as good as we hoped, but the precision is rather low. This is mostly due to the spurious RANK constraints. We note that the tables are quite short, which increases the chance of a constraint like \textit{RANK} to be true by chance; we expect less spurious constraints on larger datasets.

@@

In this section the impact on the running time is analyzed of 1)~the number of vectors; 2)~the length of the vectors; and 3)~the size of the blocks.
This analysis is accomplished by testing our system on generated spreadsheets.

Recall that our method uses a 2-phase approach where in the first phase, the superblock assignments are generated (Section~\ref{sec:algo:super} and in the second phase, the subblock assignments are generated (Section~\ref{sec:algo:subgr}).

\paragraph{Spreadsheet generation}
For our analysis, we perform multiple experiments on synthetically generated spreadsheets.

Many constraints require either numeric or discrete data, therefore, the type of cells also has an impact on the running time.
Since integers fulfill both of those typing restrictions, the generated spreadsheets contain only integer data.

One of the assumptions guiding the implementation is that most of the possible constraints are false.
To simulate this setting, the generated spreadsheets contain random data.
There will be few constraints that hold on these spreadsheets, however, random data is likely to produce vectors that fulfill the \textit{ALLDIFFERENT} constraint, which is a requirement for many other constraint templates.

In order to also analyse constraints that require their arguments to be picked from different tables, the spreadsheets are made up of two tables.
For one table the number of vectors and block size is fixed to two, while for the other table these properties are varied.
The number of vectors and the block size in the experiments are only reported for the adaptable table.
The length of the vectors can also be adapted and applies to both tables.
Unless otherwise mentioned, the adaptable table consists of one block containing all its vectors and by default the number of vectors and vector length are both~$8$.
%The running times for various configurations are plotted in Figure~\ref{fig:runtime_analysis}. % no such general pointer, specific pointers only

\paragraph{Theoretical analysis of runtime, this does not belong in experiments}\tias{move to technical section}\sam{Has to be moved yet to 'Constraints' section, explaining fail-fast}
For most constraint templates the time required to check whether a subassignment satisfies the template is linear in the size of the subassignment (the number of subblocks and their size).
For a constraint with arity~$n$, and a subassignment consisting of $n$~blocks with $m$~vectors of length~$k$, the worst case complexity becomes $O(n * m * k)$.
However, in practice, most subassignments will not satisfy the constraint template.
Since our implementation tries to account for this by failing fast\tias{explain more precisely what you mean with 'failing fast'}, the amortized time is constant in the length of the vectors~$k$, unless $k$~is very large\footnote{Specifically, if $k$ is much larger than the number of candidate subassignments} or many constraints hold in the data.
For templates requiring their arguments to be single vectors $m$~is~$1$, and the total complexity for checking a subassignment becomes~$O(n) \simeq O(1)$ (since $n$ is typically small).

\paragraph{Effect of number of vectors}
Figure~\ref{fig:runtime_analysis}a shows the results of our experiments, in which \sname was run on generated spreadsheets containing an increasing number of (equally sized) vectors.
As shown in the figure, the number of vectors in a spreadsheet can have a significant impact on the running time.

This behavior is to be expected when not many superblock assignments can be discarded in phase~1, for example, when all vectors have the same type and size.
In this case, for a constraint template that requires $n$~single vector arguments and a spreadsheet with $v$~vectors, up to $\frac{v!}{(v - n)!}$ subassignments are generated throughout the algorithm.

Typically, we expect the number of generated subassignments to be much smaller in real world spreadsheets and, therefore, the running time to be less sensitive to the number of vectors.
This is because in phase~1 entire blocks can be disqualified that do not match the signature requirements, e.g. blocks of vectors with different lengths for the product constraint.
% \tias{Better move to 4.3.2? or have there too anyway, no new facts in 'experiment' section}\samuel{?}
\samuel{=> benchmark}

% Many constraints take single vector arguments, therefore, the total number of subassignments that are generated in the second phase of the algorithm depends on the number of vectors in the spreadsheet and the arity of the constraint.

% \tias{unclear, this is about phase 2, the number to test, or the actual testing of each? explain and relate to Sec 4.3.2 what aspect you are discussing here} to be tested depends on the number of vectors in the spreadsheets and the arity of the constraint.

% For a constraint template with arity~$n$ and $v$~vectors there are $n^v$ subassignments (or $\frac{n!}{(n - v)!}$ if the arguments have to be different). \tias{Better move to 4.3.2? or have there too anyway, no new facts in 'experiment' section}
% As the number of vectors in a spreadsheet increases, the running time will quickly be dominated by constraints with a larger arity.
% This effect is noticeable in our synthetic experiments when the number of (equally sized) vectors \tias{more details about equal sized: is this a strict requirement for this to happen? I assume this is worst case and does not happen that much in practice? (the next sentences half hint at that, be more explicit)} is increased (Figure~\ref{fig:runtime_analysis}a).

\paragraph{Effect of length of vectors}
%The benchmark experiments show that the number of cells in a spreadsheet has a very limited impact on the running time.
To measure the effect that the length of the vectors has on the running time, we ran our system on spreadsheets in which the length of the vectors was gradually increased.
Figure~\ref{fig:runtime_analysis}b summarizes our results and shows that the length of the vectors has a small impact on the performance for most templates.

As discussed in \samuel{???}, our implementation attempts to detect as fast as possible if a subassignments does not satisfy a constraint template.
Our experiments satisfy the assumption that the large majority of possible subassignments do not satisfy most constraints.
Therefore, the results confirm that our implementation is able to severely limit the impact of the length of vectors.

Most of the increase in running time can be attributed to the \textit{RANK} constraint.
The implementation is not always able to detect subassignments that do not satisfy \textit{RANK} quickly and often has to fall back to analyzing an entire vector.

\paragraph{Effect of block size}
To measure the effect of changing the block size, the number of vectors is set to~$32$ and the vectors are split into varying number of blocks (up to ~$32$).
Figure~\ref{fig:runtime_analysis}c shows the running times of \sname for various settings.
As the blocks become smaller, the total running time first decreases before increasing again.
Looking at the running times for aggregate constraints and non-aggregate constraints  separately reveals that they are affected differently by the block size.
For aggregate constraints (\samuel{xxx line}) the running time decreases with the block size, while for non-aggregate constraints (\samuel{yyy line}) the running time first stays more or less constant but then increases more strongly as the size of the blocks becomes very small.

Aggregate constraints, e.g. sum or max, have an argument that takes subblocks of varying sizes.
Contrary to non aggregate constraints, which only have single vector argument, superblock assignments from phase~1 are not split into single vectors subassignments in phase~2.
Instead, for an argument with variable size, the system tries to find any subblock of the given superblock that satisfies the constraint.
Since, for a superblock of size~$m$, the number of subblocks is $O(m^2)$ \samuel{ref?}, the search for aggregate constraints will be faster if there are more but smaller blocks\footnote{Consider, for example, $v$~vectors distributed over $b$~blocks of size $v/b$, the number of subblock candidates then is $O(b * v^2 / b^2) = O(v^2 / b)$}.

Non-aggregate constraints have only single vector arguments.
Therefore, the total number of subassignments generated by the algorithm is not influenced by the block size.
By splitting a set of vectors into many blocks, however, there will be a larger number of superblocks in phase~1 and, therefore, a larger number of superblock assignments have to be generated.
Recall that in phase~1 the algorithm additionally tests block level properties for every superblock assignment.
In the generated spreadsheets these tests can only disqualify few or no superblock assignments, therefore, it will be faster to group the same vectors into larger super blocks, generate fewer superblock assignments to test and split these into single vector subassignments in phase~2.

Splitting unrelated vectors into smaller subblocks can decrease the running time for some spreadsheets by reducing the workload for aggregate constraints.
However, the two phase approach is also well suited to operate on large blocks and eliminate superblock assignments that are not compatible with constraint template signatures. \samuel{Does this conclusion sound sound?}

% \tias{Point to figure to support this (actually, move that line into 4c??); also, if this is true, you are basically saying that the CP-based first phase is worse then the second phase? That kind-of demotivates the idea of using CP in first phase... What do you really want to share with the reader?}

\subsection{Effectiveness on real spreadsheets}
In order to test our approach on real data we assembled a benchmark of spreadsheets from three sources:
1)~spreadsheets from an exercise session for teaching Excel based on~\cite{excel_book}; 2)~spreadsheets from tutorials online; and 3)~publicly available \textit{data} spreadsheets such as real-world crime statistics or financial reports (available at the same repository\footref{github-link}).

Table~\ref{tbl:category_overview} gives an overview of the spreadsheets in the different categories and summarizes the results of our experiments.
% Notice that for the data category the average number of cells is higher while the number of intended constraints and tables is lower compared to the other categories. \tias{Weird here because purpose unclear, point to this when discussing some result for which this is important}

@@

\paragraph{Q1: Effectiveness}
Our first questions, Q1, is about the inability to find intended constraints due to excessive runtime. As Table~\ref{tbl:category_overview} (last row) shows, the runtime of our method on these spreadsheets is always fast.  
The average runtime across all spreadsheets is $0.13s$.
Across categories the average running times fluctuate in the range between $0.08s$ to $0.20s$.
While most spreadsheets can be processed very quickly, running times can go up to around $0.85s$ for others.
In the previous section we analyzed the effect of various factors that can help explain runtime behavior.
Our benchmark experiments show that our algorithm is able to perform well in practice, even for larger spreadsheets∞m.

@@

Looking at the runtime per constraint template individually, the slowest by far is fuzzy lookup using $\pm 21\%$ of the running time.
It is followed by product ($\pm 7\%$), relative difference ($\pm 6\%$), running total ($\pm 5\%$), difference ($\pm 5\%$) and various row-aggregates (each $5\% - 3\%$).

Our method uses dependencies to increase efficiency by finding constraints incrementally. To illustrate the effect this can have, we ran the benchmarks using FOREIGNKEY instead of ALLDIFFERENT as base constraint to find conditional aggregates such as SUMIF. This is a stronger assumption, and a too strong in practice as not all keys have to be used in the conditions. However, the total running times per category are decreased by 6\% (exercises), 30\% (tutorials) and~3\% (data). This shows that the use of dependencies can have a strong effect on running times by sharing the computation and pruning done for one constraint with another.\sam{Check numbers}

>>>>

\begin{abstract}
Spreadsheets, CSV (comma separated value) files and other tabular data representations are in wide use today.
However, writing, maintaining and identifying good formulas for tabular data and spreadsheets can be time consuming and error-prone.
In this work, we investigate the automatic learning of constraints (formulas and relations) in raw tabular data in an unsupervised way.
We represent common spreadsheet formulas and relations through predicates and expressions whose arguments must satisfy the inherent properties of the formula/relation. The challenge is to automatically infer the set of constraints that is present in the data, without labeled examples or user feedback. 
We propose a two-stage generate and test method where the first stage uses constraint solving techniques to efficiently reduce the number of candidates in a generic way, based on the predicate signatures. 
Our approach takes inspiration from inductive logic programming, constraint learning and constraint satisfaction.
We show that we are able to accurately discover constraints in spreadsheets from various sources.
\end{abstract}

@@

% Moved up
The question that we ask in this paper is: is it possible to learn the set of constraints present in a spreadsheet, from just flat tabular spreadsheet data? The main challenge is the number of possible constraints and combinations of rows and columns that need to be tried as input to the constraints.
To answer this challenge we propose a general-purpose method and system, named \sname (from: Tabular Constraint Learner, pronounced ``tackle''), for discovering row-wise and column-wise constraints.
%
% contribs
Our contributions are as follows:
\begin{itemize}
\item we define the tabular constraint learning problem, where the goal is to find constraints that range over entire rows or columns in an unsupervised way;
\item we propose an effective 2-stage generate and test method where the first stage reasons only over properties of contiguous blocks of rows/columns, and the second stage continues to investigate individual rows and columns and their content;
\item furthermore, in the first stage we use a constraint solver to efficiently enumerate all combinations of maximally contiguous blocks compatible with the candidate constraint's argument; %, using backtracking search and forward checking;
\item experiments on different publicly available spreadsheets show that the system is able to extract constraints with high precision and recall. 
\end{itemize}

@@

This paper is organized as follows.
Section \ref{sec:related_work} discusses related work. Section \ref{sec:formalization} introduces the relevant concepts and defines the problem formally while Section \ref{sec:approach} presents our approach. This is evaluated in Section \ref{sec:evaluation} after which we show two applications of how our system could be used in Section \ref{sec:applications}, and we conclude in Section \ref{sec:conclusions}.

@@

First, it borrows techniques from logical and relational learning \cite{luc_book}, as it finds
constraints in multiple relations (or tables). Unlike typical logical and relational learning
approaches, it focuses on data in spreadsheets and constraints involving columns and rows in the sheets.
Furthermore, it derives a set of simple ``atomic'' constraints rather than a set of rules that each consist of a conjunction of literals as in clausal discovery \cite{claudien,lallouet}. Nevertheless, several ideas from logical and relational learning
proved to be useful, such as the use of a canonical form, a refinement graph, and pruning for redundancy. It also connects with the logical generalization of given formulas in a spreadsheet \cite{Isakowitz}, where one abstracts present formulas by introducing parameters in the place of particular cell references. The key difference here is that we discover constraints (formulas) in the data, we do not have or use any prior information on the formulas in a spreadsheet.

Second, there exist a number of algorithms in the constraint programming community that induce constraints from one or more examples and questions. Two well-known approaches include ModelSeeker \cite{modelseeker} and Quacq \cite{Quacq}.
The former starts from a single example in the form of a vector of values and then exhaustively looks for the most specific constraints that hold given a large constraint library. To this aim, it cleverly enumerates different de-vectorized tables; for instance, if the initial vector was of dimension $1 \times 20$, it would consider rearrangements of size $2 \times 10$, $4 \times 5$, $5 \times 4$, $\dots$ and Modelseeker would then look for both column, row  and even diagonal constraints. Key differences with our work are that we assume given spreadsheets with \textit{multiple} tables; and that ModelSeeker's constraints are restricted to integer values and constraints that are typical for the constraint programming community, e.g. aimed at combinatorial optimization and scheduling. Conacq \cite{Conacq} acquires a conjunction of constraints using techniques inspired by Mitchell's version space algorithm to process examples; its successor Quacq uses active learning to query the user regarding satisfiability. Each of the examples/queries is processed for each constraint, hence an important issue is how fast the algorithm converges to the entire solution set (e.g. for Sudoku).
The focus is again on integer variables and typical combinatorial optimisation constraints. In a similar spirit the field of equation discovery \cite{equation_discovery} focuses, guided by  heuristics based on the structure of physical equations and unit measures, on the discovery of physical formulas over numbers and times series.
However, they focus specifically on noisy physical data and formulas, e.g. a single variable can take values from a series of measurements to get a regression parameter estimate or they apply specific data transformation such as numerical calculation of time and partial derivatives. In other words, the method works with approximations and the equation structures are heavily biased towards parameter estimation, while our goal is to discover hard constraints that hold over a subset of the rows and columns.
%As the conjunction for typical constraint programming problems (such as Sudoku) can be quite large, a key problem is to cope with redundant constraints. As Modelseeker, these systems focus on the types of constraints in combinatorial optimization but unlike ModelSeeker and our approach, they do not assume the variables are organized in tabular form.

@@

    This example combines several Excel sheets based on exercises in the book ``MS Excel 2010''~\cite{excel_book}.

@@

  \caption{Constraints learned for the tables above, except 18 \textit{ALLDIFFERENT}, 2 \textit{PERMUTATION} and 5 \textit{FOREIGNKEY} constraints not shown.

@@

\subsubsection{Constraints} \label{sec:which_cons}
The constraints that are currently supported in our system are shown in Table~\ref{table:constraints}.
We included most formulas that we encountered in tutorial spreadsheets including the popular \textit{SUM} and \textit{LOOKUP} constraints\footnote{\href{https://support.office.com/en-us/article/Excel-functions-by-category-5F91F4E9-7B42-46D2-9BD1-63F26A86C0EB}{https://support.office.com/en-us/article/Excel-functions-by-category-5F91F4E9-7B42-46D2-9BD1-63F26A86C0EB}}.
We also added three structural constraints (\textit{ALLDIFFERENT}, \textit{PERMUTATION} and \textit{FOREIGNKEY}) so that they can be used in the signature of other constraint templates; they are also popular in constraint satisfaction~\cite{modelseeker}.

@@

\paragraph{Experimental setup}
\label{sec:evalualtion:method}
All spreadsheets used in the experiments are in CSV format, and for each we have ran the table extraction tool once (using a combination of the rather primitive automatic detection + visual verification that the tables are meaningful (e.g. merge split-up tables because of None values, or remove headers of all-textual tables). \\ % split of weird sentence
Group definitions are only specified if the spreadsheet contains ambiguous \textit{None} values. \tias{What does this mean, 'group definitons'?? Is this 'block detection' (Sec 4.2), if so, how do we explain this process??}

All spreadsheets also have a set of ground-truth of constraints, we call these the \textbf{intended} constraints that are expected to be (re-)discovered.
These were determined by us using the formulas of the original sheet (if present), the headers in the sheet as well as general context if need. \tias{last part too vague!! Give example of where 'context' was needed/used. If nowhere, cut that part of sentence}.
In the following experiments, we focus on spreadsheets specifically and hence only include \textit{functional} constraints, constraints that can be expressed as formulas. We ignoring structural constraints such as \textit{ALLDIFFERENT}, \textit{FOREIGNKEY} and \textit{PERMUTATION}. All constraints are stored in their canonical form. 5 spreadsheets contained rows or columns that were entirely identical, and any intended constraint involving one such vector was also added for the equivalent vectors.

Section~\ref{sec:which_cons} discusses the 32 constraints currently supported by \sname \tias{It does, right? Should also contain '32' number}. Some of the intended constraints are currently outside of the scope of \sname however, more specifically nested mathematical or nested logical formulas. We denote by \textbf{supported} constraints the subset of intended constraints that the system can find in theory.

@@

Hence, for this example we achieve a recall of 100\% (all intended constraints are discovered) and a precision of $12/18 = 67\%$. All intended constraints are supported. The recall is as good as we hoped, but the precision is rather low. This is mostly due to the spurious RANK constraints. We note that the tables are quite short, which increases the chance of a constraint like 'RANK' to be true by chance; we expect less spurious constraints on larger datasets.

@@

There are multiple factors that can impact the running time of \sname: the number of vectors, the length of the vectors, the number of vectors per block, the type of cells, the number of tables and the number of constraints that are in the data.
Moreover, the running time depends on the number of the constraint templates implemented in the system and their arity.

In this section the impact on the running time is analyzed of 1)~the number of vectors; 2)~the length of the vectors; and 3)~the size of the blocks. \tias{Why only these and not the other? Motivated/explain, or also measure their effect ; )}

Recall that our method uses a 2-phase approach where in the first phase, the superblock assignments are generated (Section ...) and in the second phase, the subblock assignments are generated (Section ...).

\paragraph{Spreadsheet generation}
We do this analysis on a set of synthetically generated spreadsheets.
The spreadsheets contains random integers\footnote{For strings and floats the results are similar but the running times are typically much lower.} and are made up of two tables.
For one table the number of vectors and block size is fixed to two, while for the other table these properties are varied.
Moreover, the length of the vectors (for both tables) can be adapted, while the default is~$8$.
%The running times for various configurations are plotted in Figure~\ref{fig:runtime_analysis}. % no such general pointer, specific pointers only

\paragraph{Theoretical analysis of runtime, this does not belong in experiments}\tias{move to technical section}
For most constraint templates the time required to check whether a subassignment satisfies the template is linear in the size of the subassignment (the number of subblocks and their size).
For a constraint with arity~$n$, and a subassignment consisting of $n$~blocks with $m$~vectors of length~$k$, the worst case complexity becomes $O(n * m * k)$.
However, in practice, most subassignments will not satisfy the constraint template.
Since our implementation tries to account for this by failing fast\tias{explain more precisely what you mean with 'failing fast'}, the amortized time is constant in the length of the vectors~$k$, unless $k$~is very large\footnote{Specifically, if $k$ is much larger than the number of candidate subassignments} or many constraints hold in the data.
For templates requiring their arguments to be single vectors $m$~is~$1$, and the total complexity for checking a subassignment becomes~$O(n) \simeq O(1)$ (since $n$ is typically small).

\paragraph{Effect of length of vectors}
%The benchmark experiments show that the number of cells in a spreadsheet has a very limited impact on the running time.
As can be seen in Figure~\ref{fig:runtime_analysis}b\tias{This is the first figure discussed so must be a}, increasing the length of the vectors has a negligible impact on the performance for most template. Most of the increase in running time can be attributed to the RANK constraint which is less suited for failing fast \tias{Too vague, 'less suited', be precise}; the runtime for detecting all constraints except RANK is shown in green to highlight this.

\paragraph{Effect of number of vectors}
On the contrary, the number of vectors in a spreadsheet can have a significant impact on the running time.
Many constraints take single vector arguments, therefore, the number of subassignments \tias{unclear, this is about phase 2, the number to test, or the actual testing of each? explain and relate to Sec 4.3.2 what aspect you are discussing here} to be tested depends on the number of vectors in the spreadsheets and the arity of the constraint.
For a constraint template with arity~$n$ and $v$~vectors there are $n^v$ subassignments (or $\frac{n!}{(n - v)!}$ if the arguments have to be different). \tias{Better move to 4.3.2? or have there too anyway, no new facts in 'experiment' section}
As the number of vectors in a spreadsheet increases, the running time will quickly be dominated by constraints with a larger arity.
This effect is noticeable in our synthetic experiments when the number of (equally sized) vectors \tias{more details about equal sized: is this a strict requirement for this to happen? I assume this is worst case and does not happen that much in practice? (the next sentences half hint at that, be more explicit)} is increased (Figure~\ref{fig:runtime_analysis}a).
In practice, in our benchmark experiments, the assignment step of our algorithm is able to limit this rapid increase in running time for larger numbers of vectors \tias{How? explain, if you explain in the next sentence then start it with 'This is because' or alike, reads better}.
The superblock assignment phase is able to disqualify entire blocks that do not match the signature requirements, i.e. blocks of vectors with different lengths for the product constraint.

Aggregate constraints, e.g. sum or max, have an argument that takes subblocks of varying sizes.
Therefore, the size of the blocks \tias{in what phase? the max. contigiuous? be more precise} in the spreadsheet also influences the runtime of these constraints.
For a block of size~$m$ there are $\frac{m(m + 1)}{2}$ or $O(m^2)$ subblocks, thus, if a spreadsheet has more but smaller sized blocks, the search for aggregate constraints will be faster.
Consider, for example, $v$~vectors distributed over $b$~blocks of size $v/b$, the number of subblock candidates then is $O(b * v^2 / b^2) = O(v^2 / b)$.
The \tias{explain what THE experiment is, and how you set it up} synthetic experiments confirm that splitting a constant number of vectors into more, smaller blocks improves the performance of the aggregate constraint search (Figure~\ref{fig:runtime_analysis}c, green line).

These experiments, however, also show that the running time for various non-aggregate constraints increases as the size of blocks becomes small.
Since these constraints have single vector arguments, the block size should not have a strong effect.
However, in practice, when there are many small blocks, the superblock assignment phase has to generate a very large number of combinations, while for larger blocks this work is done mostly during the subassignment generation phase.
Since the implementation of that last phase is more adept at generating these combinations and does not need to check for constraints on all these combinations, shifting this workload to the assignment \tias{more precise: to superblock?} phase slows down the algorithm. \tias{Point to figure to support this (actually, move that line into 4c??); also, if this is true, you are basically saying that the CP-based first phase is worse then the second phase? That kind-of demotivates the idea of using CP in first phase... What do you really want to share with the reader?}

\subsection{Effectiveness on real spreadsheets}
In order to obtain a benchmark of spreadsheets we looked at three sources:
1)~spreadsheets from an exercise session for teaching Excel based on~\cite{excel_book}; 2)~spreadsheets from tutorials online; and 3)~publicly available \textit{data} spreadsheets such as real-world crime statistics or financial reports (available at the same repository\footref{github-link}).

Table~\ref{tbl:category_overview} gives an overview of the spreadsheets in the different categories and summarizes the results of our experiments.
Notice that for the data category the average number of cells is higher while the number of intended constraints and tables is lower compared to the other categories. \tias{Weird here because purpose unclear, point to this when discussing some result for which this is important}

@@

\paragraph{Q1: Effectiveness}
Our first questions, Q1, is about the inability to find intended constraints due to excessive runtime. As Table~\ref{tbl:category_overview} (last row) shows, the runtime of our method on these spreadsheets is always fast.  
The average runtime across all spreadsheets is $0.13s$.
Across categories the average running times fluctuate in the range between $0.09s$ to $0.22s$.
While most spreadsheets can be analyzed very quickly, running times can go up to around $1s$ for others. ... \tias{Explain by relating to previous section of synthetic data}

@@

Looking at the runtime per constraint template individually (including equality\tias{what does this mean?}), the slowest by far is fuzzy lookup using $\pm 36\%$ of the running time.
It is followed by product ($\pm 5\%$), relative difference ($\pm 5\%$) and various row-aggregates (each $3.5\% - 4\%$).

Our method uses dependencies to increase efficiency by finding constraints incrementally. To illustrate the effect this can have, we ran the benchmarks using FOREIGNKEY instead of ALLDIFFERENT as base constraint to find conditional aggregates such as SUMIF. This is a stronger assumption, and a too strong in practice as not all keys have to be used in the conditions. However, the total running times per category are decreased by 6\% (exercises), 30\% (tutorials) and~3\% (data). This shows that the use of dependencies can have a strong effect on running times by sharing the computation and pruning done for one constraint with another.
